Four-body L=1 systems from contact EFT

### 2022-01-06 manuscript/Draft_0.2.tex
DC initial commit

\begin{abstract} 
(to be cleaned at the end)
Few-body scattering resonances appear in multiple physical fields and are commonly found in experiments. 
Compared with boundstates, they are more difficult to be studied theoretically and numerically because of their belonging to the continuum phase-space.
In this study, we analyze the minimal theory that predicts the appearance of $L=1$ four-body resonance.
This state is known in nuclear physics to be related to the lighter nucleus that shows resonant behavior: $^4$H.
However, we aim not only to study it in the nuclear framework but also in more general universal systems.
For this purpose, we employ a contact effective field theory both fitted on nuclear observables (pionless effective field theory) and to unitary systems (universal effective field theory).
This study will help to understand how resonances emerge in few- and many-body systems for nuclear physics and any other physical field close to universality.
\end{abstract}

@@

Resonances are ever-present in quantum and classical physics.
However, especially in many-particle quantum fields as hadronic, nuclear, and atomic physics, they are far from been trivial to be predicted and calculated. 
A resonance in these fields is generally a consequence of a nonperturbative and complex interplay of particle-particle interactions and open channels that leads to the existence of multiple excitation and thresholds. 
The broad size and many-body nature make these states also relatively hard to be identified experimentally. 
Although many models predict the existence of resonant states with better or worse success, just looking at the Hamiltonian of a system it is hard to foresee their existence, and the numerical methods that can extract complex energy are, in general, expensive or less accurate than the ones developed for bound states.
%

%
Among nuclear systems with resonances, $^4$H $J^\pi=1^-$ has a special role and has been object of several studies as it is known to be the smallest nucleus with a (relatively broad) resonance.
%
In particular, cross-sections and scattering parameters have been studied extensively experimentally (e.g. \cite{osti_4230875, Phillips:1980zz, Tilley:1992zz}) and theoretically.
Concerning the theoretical side, several interactions and different numerical methods have been tested
(see, e.g. \cite{Ciesielski:1997vy,Ciesielski:1998sy,Ciesielski:1999pp,Viviani:1998gr,Fonseca:1999zz,Lazauskas:2004uq}) to predict well phaseshifts especially for low-energy scattering.
%
Fewer studies can be found, instead, about the position of the $^4$H resonance pole that induces such phaseshift. 
In \cite{Arai:2003ek,deDiego:2007rd,Lazauskas:2019cxj} the resonance position of $^4$H is extracted using respectively, RGM (which will be explained in the next sections of this paper), two-body $n-t$ potential, and three microscopic potentials with a Faddeev-Yakubovsky solver. 
The theoretical models agree for the presence of a $J^\pi=1^-$ resonance pole with energy $Re(E)=\{0.90 - 1.23\}$ MeV and width $\Gamma=\{3.5 - 5.8\}$ MeV.
Experimentally, the resonant pole can be extracted by R-matrix analysis to be at $Re(E)=3.50$ MeV and  $\Gamma=6.73$ MeV \cite{Tilley:1992zz}.
Although there is no precise agreement among experiment and models on the energy position of this resonant pole, the ability to create it in such models is still remarkable and has intrinsic value in the interpretation of the experimental cross-section.
%
The importance of studying $^4$H resonances and of the theoretical mechanisms that create it exceed the description of the system itself. 
It is the getaway for larger systems: on one hand for many-neutrons, on the direction of studying the stability of neutron tears and eventually to the neutron star equation of state.
On the other hand, to the study of resonances in heavy nuclei that are nowadays still hard or even impossible from ab-initio principles. 
%
In this study, we are interested in both these aspects on the framework of very minimalists nuclear forces: renormalizable contact effective field theories. 
%

%
In the last years, effective field theories (EFT) have had an ever-increasing impact on nuclear and atomic physics. 
They rely on the expansion of the appropriate underline theory onto a set of operators hierarchically arranged in a powercounting. 
The origin of this expansion depends on the kind of system in study and of the interparticle typical momentum exchanged, and differentiate an EFT from another. 
For example, it was shown that few-body nuclear systems are close to the unitary limit, in which the two-body scattering length is much larger than any other scale in such systems, allowing the description of few nucleons expanding the EFT around an infinite (or large) scattering length (see \cite{vanKolck:1999mw}).
This takes the name of pionless powercounting (\eftnopi) and, contrary to the chiral EFT, in which pions origin long-range forces, it is an expansion in contact operators.
The absence of pions in \eftnopi limits this theory to be a low-momentum theory (i.e. can not describe phenomena which typical momentum would allow the creation of pions) but allows it to be fully renormalizable and relatively easy to be treated numerically and analytically.
This makes it best suited to understand in simple terms even complex many-body phenomena and for the description of very shallow quantum states, like resonant poles. 
In the last years, \eftnopi was proven to be a valuable theory to study the behavior of many-boson systems and nuclei up to four nucleons. 
However, the LO of the theory does not appear to perform as well for the behavior of multi-fermionic systems, on which it erroneously predicts unbound nuclei above four particles (eg. $^7$Li and $^{16}$O) \cite{Schafer:2020ivj,Contessi:2017rww}.
The reason for this is to be attributed to the theory's more general inability of creating bound mixed symmetry states.
However, the subleading order of the EFT may still be able to correct this behavior provided that the LO creates, at least, an unbound state (virtual o resonant) that can be moved to the correct bound position. 
On the contrary, if no state is present at LO in these systems, of which $^4$H is the minimal representative, only a modification of the powercounting, and therefore of the entire theory, will allow for the correct description of nuclear physics using contact EFTs.
The creation of resonant poles in LO \eftnopi is possible, as it was shown in $nn\Lambda$ systems \cite{Schafer:2020rba} (as a subthreshold resonance), however, there is no record of such poles created in systems with mixed symmetry groundstates as $^4$H. 
By approaching the existence of a resonant pole in $^4$H we are tackling two problems at the same time: the possibility of creating resonances with the barest possible interaction and the fact that contact EFT can or not describe nuclear physics.
%

%
To access this state we first fix a \eftnopi on nuclear observables (fixing the nucleon mass, and the deuterium and triton bindings).
The same theory was also fixed to nuclear systems in which the two-body potential was tuned to the unitary limit.
This last case allows access to the universal limit and the deviation of the nuclear potential from it.
We chose to use a SU(4) symmetric theory in both cases for simplicity as this was proven to be a good first-order representation of nuclear physics \cite{Konig:2016utl}.

@@

>>>>

\begin{abstract} 
Scattering resonances are ubiquitous in physics, and for the understanding of the behaviour of few-body quantum systems as important as bound states.
Compared with the latter, their theoretical handling and their role in
the interpretation of data is complicated through their unstable nature
which is reflected in their wave functions not being elements of a
square-integrable space.
The smallest nuclear system comprised solely of resonances
is 4-hydrogen with $L=1$ states about $3.5$~MeV above the triton-neutron
threshold.

In this article, we analyze the minimal contact theory which predicts
such $P$-wave, four-body resonances. Furthermore, we establish numerically
the emergence of this specific resonance pole for a wider class of systems
whose two-body scale is significantly larger ($\to\infty$) compared with the
interaction range.
Specifically, we employ a contact effective field theory renormalized
with nuclear observables -- in which case the interaction is identical
to the leading order of the pionless effective field theory --
and to unitary systems without a finite two-body scale (universal
contact effective field theory). We find a negative-parity four-body
resonance pole which is stable with respect to the short-distance structure
of the regularized two- and three-body momentum-independent contact
potentials with three independent numerical algorithms: a finite-volume
Gaussian-stochastic-variational technique, the solution of the AGS on a
discrete coordinate-space grid, and a one-channel, two-body,
resonating-group reduction of the problem.
Thereby we demonstrate that a zero-range potential
provides a non-trivial pole structure for the four-body amplitude.
\end{abstract}

@@

Resonance phenomena occur ubiquitously in quantum and classical physics.
Relative to stable bound states, their handling in many-particle hadronic, nuclear,
and atomic physics is complicated. 
A resonance results from a nonperturbative interplay of particle-particle interactions,
fine-tuned potential shapes, and channel couplings that lead to multiple
excitations and thresholds.
Relative long life times and energies very close to production thresholds lead to
clean, measurable signals, while in practise, a rapid decay and an energy far from
threshold makes their identification with cross section data involved. 
Theoretically, the structure of a Hamiltonian usually does not expose the existence of
resonances without explicitly solving dynamical equations. In the latter course,
numerical methods to extract complex energies are, in general, expensive and
less accurate compared with those for bound states.
%

%
In order to identify general features of an interaction which allow for the
prediction of resonances, we focus on the smallest nuclear system, known to
be characterized by a resonance, namely, the 4-hydrogen ($^4$H).
In this system, $J^\pi=0^-, 1^-, 2^-$ resonances are found experimentally approcimately
$5.2$~MeV, $3.5$~MeV, and $3.2$~MeV, respectively, above the triton-neutron threshold.
Cross sections and scattering parameters have been studied extensively both experimentally
~\cite{osti_4230875, Phillips:1980zz, Tilley:1992zz}~and theoretically~
\cite{Ciesielski:1997vy,Ciesielski:1998sy,Ciesielski:1999pp,Viviani:1998gr,Fonseca:1999zz,Lazauskas:2004uq}.
The position of the $^4$H, $J^\pi=1^-$ resonance pole was located in
Refs.~\cite{Arai:2003ek,deDiego:2007rd,Lazauskas:2019cxj}~ at an energy
$Re(E)=\{0.90 - 1.23\}$~MeV and width $\Gamma=\{3.5 - 5.8\}$~MeV.
From data, a $R$-matrix analysis puts this pole at $Re(E)=3.50$~MeV and $\Gamma=6.73$~MeV~
\cite{Tilley:1992zz}.
Despite this discrepancy, the sheer existence of the pole is remarkable.
%
%
%
Why does
such a pole emerge in the four-body amplitude but is absent, e.g., in the
deuteron-neutron quartet channel or the three-neutron system? What are minimal
conditions and characteristics of the employed models necessary for these features?
These questions are crucial in the development of the nuclear effective field
theory in order to make it useful for the description of heavier nuclei beyond
the $S$ shell.
Furthermore, fermi statistics which enforces higher relative angular
momenta in ground-state wave functions of more than four nucleons, is equally
relevant for hadronic, atomic, in general, all fermions amenable
to a description with zero-range forces. Hence, an understanding how the structure
of these forces leads to certain few-body resonance poles carries value far beyond the
$^4$H system, namely: is a leading-order description of a system of particles, whose
statistics demand mixed-orbital-symmetry states, with properly renormalized coupling
strengths of zero-range, possibly spin-/flavour-dependent interactions possible?
%In this study, we are interested in both these aspects on the framework of very minimalists nuclear forces: renormalizable contact effective field theories. 
%
%
%
%

%
Such renormalizable effective field theories (EFT) were instrumental in understanding
the relation between nuclear physics and its underlying theory of quantum chromodynamics
and for deriving universal features which nuclei share with atoms.
All EFT's comprise of the most general Lagrangean whose terms are built with
appropriately chosen degrees of freedom and meet certain internal and space-time symmetries.
A hierarchy (power counting) amongst this infinity of operators and their weight within
an iterative expansion of the $S$-matrix renders the theory {\it effective}, i.e., useful.
In the two-body sector, the convergence of the expansion and the choice of the relevant
degrees of freedom depends on the typical momentum exchanged between interaction partners;
for different typical momenta, different EFT expansions are needed. 
For instance, few-body nuclear systems are close to the unitary limit, where the two-body
scattering length is much larger than any other scale, an
expansion around an infinite (or large) scattering length with solely nuclear degrees of freedom
yields the bulk structure of nuclei up to 4-helium~\cite{vanKolck:1999mw}.
For momenta that allow for the creation of pions, this pion-less power counting (\eftnopi)
fails. For these momenta, chiral EFTs couple the pion field to the nucleons and thereby
introduce long-range forces at leading order, while \eftnopi~is of zero-range at
leading order and thus easier to treat numerically and analytically.
Whether this simplicity of a zero-range theory still does not lead to an exclusion
of complex many-body phenomena, e.g., shallow quantum states and resonance poles is
unknown and shall be studied in this article.

What is known about \eftnopi, in particular, is its ability to predict the behavior
of many-boson systems and nuclei up to four nucleons. 
However, the leading order (LO) of the theory cannot account for the observed behavior of
multi-fermionic systems. There, it erroneously predicts unbound nuclei above four
particles, e.g.~$^7$Li and $^{16}$O~\cite{Schafer:2020ivj,Contessi:2017rww}.
It was conjectured more generally, that zero-range theories cannot sustain 
bound states of mixed symmetry.
However, the bound character of a state might appear only at a higher orders of the EFT
provided that the LO creates a pole of other kind, virtual or resonant, which the
modification of the interaction at higher order transforms into a bound state. 
If no such pole is present at LO in these systems -- of which $^4$H is the smallest --
only a modification of the power counting
will furnish the correct description of nuclear physics with contact EFTs.

The creation of resonant $S$-wave poles at LO \eftnopi~was demonstrated in case of the
hyper triton ($nn\Lambda$) in Ref.~\cite{Schafer:2020rba} (as a subthreshold resonance) but
there is no record of such poles being created renormalization-scheme independently
in systems with mixed symmetry ground states
such as $^4$H with a zero-range theory. 
It is the aim of this work to analyze this creation of a four-body $P$-wave resonance
pole from an zero-range, spherically-symmetric regularized contact interaction with
coupling strengths resembling nuclear systems close to unitarity and also systems which
realize this limit with an infinite two-body scattering length, i.e., zero-energy bound
state.

In the following section, we introduce the Hamiltonian of the theory along with
comments on its solution with a Gaussian-stochastic-variational method, the
treatment of the scattering problem via the Alt, Grassberger, and Sandhas equations
in coordinate space, and the reduction of $N$-body, two-fragment scattering onto a
two-body problem.
Following this, we present results for effective-range parameters and pole positions
for the low-energy scattering of a neutron on $^3$H before generalizing this to a
generic point particle impinging on a universal trimer.
Finally, we relate the regulator independence of the results to the deployed minimal
theory as a first-order approximation of nuclei and universal, unitary systems.

@@

The minimal theory for the

### 2022-01-06 manuscript/Draft_0.3.tex
DC initial commit

\begin{abstract} 
Scattering resonances are ubiquitous in physics, and for the understanding of the behaviour of few-body quantum systems as important as bound states.
Compared with the latter, their theoretical handling and their role in
the interpretation of data is complicated through their unstable nature
which is reflected in their wave functions not being elements of a
square-integrable space.
The smallest nuclear system comprised solely of resonances
is 4-hydrogen with $L=1$ states about $3.5$~MeV above the triton-neutron
threshold.

In this article, we analyze the minimal contact theory which predicts
such $P$-wave, four-body resonances. Furthermore, we establish numerically
the emergence of this specific resonance pole for a wider class of systems
whose two-body scale is significantly larger ($\to\infty$) compared with the
interaction range.
Specifically, we employ a contact effective field theory renormalized
with nuclear observables -- in which case the interaction is identical
to the leading order of the pionless effective field theory --
and to unitary systems without a finite two-body scale (universal
contact effective field theory). We find a negative-parity four-body
resonance pole which is stable with respect to the short-distance structure
of the regularized two- and three-body momentum-independent contact
potentials with three independent numerical algorithms: a finite-volume
Gaussian-stochastic-variational technique, the solution of the AGS on a
discrete coordinate-space grid, and a one-channel, two-body,
resonating-group reduction of the problem.
Thereby we demonstrate that a zero-range potential
provides a non-trivial pole structure for the four-body amplitude.
\end{abstract}

@@

%
In order to identify general features of an interaction which allow for the
prediction of resonances, we focus on the smallest nuclear system, known to
be characterized by a resonance, namely, the 4-hydrogen ($^4$H).
In this system, $J^\pi=0^-, 1^-, 2^-$ resonances are found experimentally approcimately
$5.2$~MeV, $3.5$~MeV, and $3.2$~MeV, respectively, above the triton-neutron threshold.
Cross sections and scattering parameters have been studied extensively both experimentally
~\cite{osti_4230875, Phillips:1980zz, Tilley:1992zz}~and theoretically~
\cite{Ciesielski:1997vy,Ciesielski:1998sy,Ciesielski:1999pp,Viviani:1998gr,Fonseca:1999zz,Lazauskas:2004uq}.
The position of the $^4$H, $J^\pi=1^-$ resonance pole was located in
Refs.~\cite{Arai:2003ek,deDiego:2007rd,Lazauskas:2019cxj}~ at an energy
$Re(E)=\{0.90 - 1.23\}$~MeV and width $\Gamma=\{3.5 - 5.8\}$~MeV.
From data, a $R$-matrix analysis puts this pole at $Re(E)=3.50$~MeV and $\Gamma=6.73$~MeV~
\cite{Tilley:1992zz}.
Despite this discrepancy, the sheer existence of the pole is remarkable.
%
%
%
Why does
such a pole emerge in the four-body amplitude but is absent, e.g., in the
deuteron-neutron quartet channel or the three-neutron system? What are minimal
conditions and characteristics of the employed models necessary for these features?
These questions are crucial in the development of the nuclear effective field
theory in order to make it useful for the description of heavier nuclei beyond
the $S$ shell.
Furthermore, fermi statistics which enforces higher relative angular
momenta in ground-state wave functions of more than four nucleons, is equally
relevant for hadronic, atomic, in general, all fermions amenable
to a description with zero-range forces. Hence, an understanding how the structure
of these forces leads to certain few-body resonance poles carries value far beyond the
$^4$H system, namely: is a leading-order description of a system of particles, whose
statistics demand mixed-orbital-symmetry states, with properly renormalized coupling
strengths of zero-range, possibly spin-/flavour-dependent interactions possible?
%In this study, we are interested in both these aspects on the framework of very minimalists nuclear forces: renormalizable contact effective field theories. 
%
%
%
%

@@

\section{Appendix: Resonating group method and interactions}
\end{document}

>>>>

\begin{abstract} 
Scattering resonances are ubiquitous in physics, and for the understanding of the behaviour of few-body quantum systems as important as bound states.
Compared with the latter, their theoretical handling and their role in the interpretation of data is complicated through their unstable nature which is reflected in their belonging to the continuum spectrum.
The smallest nuclear system comprised solely of resonances is 4-hydrogen with $L=1$ states about $3.5$~MeV above the triton-neutron
threshold.

In this article, we analyze the minimal contact theory which predicts
such $P$-wave, four-body resonances. Furthermore, we establish numerically
the emergence of this specific resonance pole for a wider class of systems
whose two-body scale is significantly larger ($\to\infty$) than their interaction range.
Specifically, we employ a contact effective field theory renormalized
with nuclear observables (LO pionless effective field theory)
and to unitary systems without a finite two-body scale (universal
contact effective field theory). We find a negative-parity four-body
resonance pole which is stable with respect to the short-distance structure
of the regularized two- and three-body momentum-independent contact
potentials with three independent numerical algorithms: a finite-volume
Gaussian-stochastic-variational technique, the solution of the AGS on a
discrete coordinate-space grid, and a one-channel, two-body,
resonating-group reduction of the problem.
Thereby we demonstrate that a zero-range potential
provides a non-trivial pole structure for the four-body amplitude.
\end{abstract}

@@

%
In order to identify general features of an interaction which allow for the
prediction of resonances, we focus on  the 4-hydrogen ($^4$H), the smallest nuclear system known to
be characterized by a resonance.
In this system, $J^\pi=0^-, 1^-, 2^-$ resonances are found experimentally approximately
$5.2$~MeV, $3.5$~MeV, and $3.2$~MeV, respectively, above the triton-neutron threshold.
Cross sections and scattering parameters have been studied extensively both experimentally
~\cite{osti_4230875, Phillips:1980zz, Tilley:1992zz}~and theoretically~
\cite{Ciesielski:1997vy,Ciesielski:1998sy,Ciesielski:1999pp,Viviani:1998gr,Fonseca:1999zz,Lazauskas:2004uq}.
The position of the $^4$H, $J^\pi=1^-$ resonance pole was located in
Refs.~\cite{Arai:2003ek,deDiego:2007rd,Lazauskas:2019cxj}~ at an energy
$Re(E)=\{0.90 - 1.23\}$~MeV and width $\Gamma=\{3.5 - 5.8\}$~MeV.
From data, a $R$-matrix analysis puts this pole at $Re(E)=3.50$~MeV and $\Gamma=6.73$~MeV~
\cite{Tilley:1992zz}.
Despite the quantitative discrepancy among theory and experiment, the sheer existence of the pole in different models is remarkable.
%
%
%
Why does
such a pole emerge in the four-body amplitude but is absent, e.g., in the
deuteron-neutron quartet channel or the three-neutron system? What are minimal
conditions and characteristics of the employed models necessary for these features?
These questions are crucial in the development of the nuclear effective field
theory in order to make it useful for the description of heavier nuclei beyond
the $S$ shell.
Furthermore, fermi statistics which enforces higher relative angular
momenta in ground-state wave functions of more than four nucleons, is equally
relevant for hadronic, atomic, in general, all fermions amenable
to a description with zero-range forces. Hence, an understanding how the structure
of these forces leads to certain few-body resonance poles carries value far beyond the
$^4$H system, namely: is a leading-order description of a system of particles, whose
statistics demand mixed-orbital-symmetry states, with properly renormalized coupling
strengths of zero-range, possibly spin-/flavour-dependent interactions possible?
%In this study, we are interested in both these aspects on the framework of very minimalists nuclear forces: renormalizable contact effective field theories. 
%
%
%
%

@@

\section{Appendix: Resonating group method and interactions}
In this section, we reformulate the $A$-body problem of one particle scattering elastically off
an $(A-1)$-body bound state with a relative kinetic energy which is small relative to the lowest
excitation energy of the target compound.

\end{document}
